---
title: "Practical Statistics for Data Scientists, Chapter 1: Exploratory Data Analysis"
author: "Original Code: Bruce, Peter C., Andrew Bruce and Peter Gedeck | Modifications: Antti Rask"
date: "2023-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 Exploratory Data Analysis

```{r}
library(conflicted)
    conflicts_prefer(dplyr::filter)
library(corrr)
library(crosstable)
library(DT)
library(gmodels)
library(matrixStats)
library(scales)
library(tidyverse)
```

## Import the Datasets Needed

```{r}
state         <- read_csv("../data/state.csv")
dfw           <- read_csv("../data/dfw_airline.csv")
sp500_px      <- read_csv("../data/sp500_data.csv.gz") %>% 
    select(date = ...1, everything())
sp500_sym     <- read_csv("../data/sp500_sectors.csv")
kc_tax        <- read_csv("../data/kc_tax.csv.gz")
lc_loans      <- read_csv("../data/lc_loans.csv")
airline_stats <- read_csv("../data/airline_stats.csv") %>%
    mutate(airline = factor(airline, levels = c(
        "Alaska", "American", "Jet Blue", "Delta", "United", "Southwest"
    )))
```

## Estimates of Location

### Example: Location Estimates of Population and Murder Rates

```{r}
# Table 1-2
state %>%
    mutate(Population = Population %>% number(big.mark = ",")) %>%
    slice_head(n = 8) %>%
    datatable(colnames = c("State", "Population", "Murder rate", "Abbreviation"))
```

```{r}
population <- state %>% pull(Population) 

population %>% mean()
population %>% mean(trim = 0.1)
population %>% median()

# Using {matrixStats}
murder_rate <- state %>% pull(Murder.Rate) 

murder_rate %>% weightedMean(w = state$Population)
murder_rate %>% weightedMedian(w = state$Population)
```

## Estimates of Variability

```{r}
# Table 1-3
state %>% 
    mutate(Population = Population %>% number(big.mark = ",")) %>%
    slice_head(n = 8) %>%
    datatable(colnames = c("State", "Population", "Murder rate", "Abbreviation"))
```

```{r}
population %>% sd()
population %>% IQR()
population %>% mad()
```

## Percentiles and Boxplots

```{r}
murder_rate %>%
    quantile(p = c(0.05, 0.25, 0.5, 0.75, 0.95)) %>%
    as_tibble(rownames = "percentile")

state %>% 
    ggplot(aes(Population / 1000000)) +
    geom_boxplot() +
    coord_flip() +
    theme_classic() +
    labs(x = "Population (millions)")
```

## Frequency Table and Histograms

```{r}
cut_interval(
    population,
    n      = 10,
    closed = c("left")
) %>% 
    table() %>% 
    as_tibble(rownames = "bin_number") %>% 
    rename(
        bin_range = ".",
        count     = n
    )

# If we just want to add the bin_range to the state tibble (we do lose the empty bins!):
state %>% 
    mutate(
        bin_range = cut_interval(
            population,
            n      = 10,
            closed = c("left")
        )
    ) %>% 
    arrange(desc(Population))
```

```{r}
state %>% 
    ggplot(aes(Population)) +
    geom_histogram(
        breaks = seq(
            from   = min(state$Population),
            to     = max(state$Population),
            length = 11
        ),
        color    = "black",
        fill     = "white"
    ) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_classic()
```

## Density Estimates

```{r}
# Density is an alternative to histograms that can provide more insight into the distribution of the data points.
state %>%
    ggplot(aes(Murder.Rate)) +
    geom_histogram(
        aes(y = after_stat(density)),
        bins  = 12,
        color = "black",
        fill  = "white"
    ) +
    geom_density(
        color      = "blue",
        linewidth  = 1.5
    ) +
    scale_x_continuous(
        breaks = seq(0, 10, by = 2),
        expand = c(0, 0)
    ) +
    scale_y_continuous(
        expand = c(0, 0)
    ) +
    labs(
        x = "Murder rate (per 100,000)",
        y = "Density"
    ) +
    theme_classic()
```

## Exploring Binary and Categorical Data

```{r}
dfw %>%
    pivot_longer(cols = everything()) %>% 
    ggplot(aes(fct_reorder(name, value), value)) +
    geom_col(
        color = "black",
        fill  = "white"
    ) +
    coord_flip() +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
        x = "Cause of Delay",
        y = "Count"
    ) +
    theme_classic()
```

## Correlation

```{r}
telecom_vec <- sp500_sym %>%
    filter(sector == "telecommunications_services") %>%
    pull(symbol)

telecom <- sp500_px %>%
    filter(date > "2012-07-01") %>% 
    select(
        date,
        all_of(telecom_vec)
    )

telecom_cor <- telecom %>%
    correlate()

telecom_cor

# Next we focus on funds traded on major exchanges (sector == 'etf')
etfs_vec <- sp500_sym %>%
    filter(sector == "etf") %>%
    pull(symbol)

etfs <- sp500_px %>%
    filter(
        date > "2012-07-01"
    ) %>% 
    select(
        date,
        all_of(etfs_vec)
    )

etfs %>%
    correlate() %>% 
    rplot(print_cor = FALSE)
```

## Scatterplots

```{r}
telecom %>%
    ggplot(aes(T, VZ)) +
    geom_point(alpha = 0.3) +
    geom_vline(
        xintercept = 0,
        color      = "grey"
    ) +
    geom_hline(
        yintercept = 0,
        color      = "grey"
    ) +
    labs(
        x = "ATT (T)",
        y = "Verizon (VZ)"
    ) +
    theme_classic()

dim(telecom)
```

## Exploring Two or More Variables

```{r}
# Filter based on a variety of criteria
kc_tax0 <- kc_tax %>% 
    filter(
        TaxAssessedValue < 750000 &
        SqFtTotLiving    > 100 &
        SqFtTotLiving    < 3500
    )

nrow(kc_tax0)
```

## Hexagonal Binning and Contours

```{r}
# Plotting numeric versus numeric data
graph <- kc_tax0 %>% 
    ggplot(aes(SqFtTotLiving, TaxAssessedValue)) + 
    stat_binhex(color = "white") +
    scale_fill_gradient(
        low  = "white",
        high = "blue"
    ) +
    labs(
        x = "Finished Square Feet",
        y = "Tax-Assessed Value"
    ) +
    theme_classic()

graph
```

## Visualize as a Two-Dimensional Extension of the Density Plot

```{r}
graph_2 <- kc_tax0 %>% 
    ggplot(aes(SqFtTotLiving, TaxAssessedValue)) +
    geom_point(
        color = "blue",
        alpha = 0.1
    ) +
    geom_density2d(color = "white") +
    labs(
        x = "Finished Square Feet",
        y = "Tax-Assessed Value"
    ) +
    theme_classic()

graph_2
```

## Two Categorical Variables

```{r}
CrossTable(
    lc_loans %>% pull(grade),
    lc_loans %>% pull(status)
)
```

```{r, warning=FALSE}
lc_loans %>%
    crosstable(
        cols            = grade,
        by              = status,
        label           = FALSE,
        total           = "both",
        percent_pattern = "{n} ({p_row}/{p_col})",
        percent_digits  = 1
    ) %>%
    as_flextable(compact = TRUE)
```

## Categorical and Numeric Data

```{r}
# Boxplots of a column can be grouped by a different column.
airline_stats %>% 
    ggplot(aes(pct_carrier_delay, airline), ylim = c(0, 50)) +
    geom_boxplot() +
    coord_flip() +
    labs(
        x = "Daily % of Delayed Flights",
        y = "Airline"
    ) +
    theme_classic()
```

```{r}
# Variation of boxplots called _violinplot_.
graph_3 <- airline_stats %>% 
    ggplot(aes(airline, pct_carrier_delay)) +
    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), linetype = 2) +
    geom_violin(fill = NA, linewidth = 1.1) +
    coord_cartesian(ylim = c(0, 50)) +
    labs(
        x = "",
        y = "Daily % of Delayed Flights"
    ) +
    theme_classic()

graph_3
```

## Visualizing Multiple Variables

```{r}
graph_4 <- kc_tax0 %>%
    filter(ZipCode %in% c(98188, 98105, 98108, 98126)) %>%
    ggplot(aes(SqFtTotLiving, TaxAssessedValue)) +
    stat_binhex(color = "white") +
    facet_wrap(vars(ZipCode)) +
    scale_fill_gradient(low = "gray95", high = "black") +
    labs(
        x = "Finished Square Feet",
        y = "Tax-Assessed Value"
    ) +
    theme_classic()

graph_4
```
