---
title: "Practical Statistics for Data Scientists, Chapter 5: Classification"
author: "Original Code: Bruce, Peter C., Andrew Bruce and Peter Gedeck | Modifications: Antti Rask"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 5 Classification

```{r}
library(conflicted)
    conflicts_prefer(dplyr::filter)
    conflicts_prefer(dplyr::select)
library(FNN)
library(klaR)
library(MASS)
library(mgcv)
library(rpart)
library(tidyverse)
```

## Import the Datasets Needed

```{r}
loan3000       <- read_csv("../data/loan3000.csv") %>%
    mutate(across(where(is.character), as.factor))

loan_data      <- read_csv("../data/loan_data.csv.gz") %>%
    mutate(across(where(is.character), as.factor))

full_train_set <- read_csv("../data/full_train_set.csv.gz") %>%
    mutate(across(where(is.character), as.factor))

# Order the outcome variable
loan3000 <- loan3000 %>%
    mutate(outcome = factor(outcome, levels=c("paid off", "default")))

loan_data <- loan_data %>%
    mutate(outcome = factor(outcome, levels=c("paid off", "default")))

full_train_set <- full_train_set %>%
    mutate(outcome = factor(outcome, levels=c("paid off", "default")))
```

## Naive Bayes

### The Naive Solution

```{r}
naive_model <- NaiveBayes(
    outcome ~ purpose_ + home_ + emp_len_,
    data = na.omit(loan_data)
)

naive_model$table

new_loan <- loan_data %>%
    filter(
        `...1` == 147
    ) %>%
    select(
        purpose_,
        home_,
        emp_len_
    )

predict(naive_model, new_loan)
```

#### Example not in book

```{r}
less_naive <- NaiveBayes(
    outcome ~
        borrower_score +
        payment_inc_ratio +
        purpose_ +
        home_ +
        emp_len_,
    data = loan_data
)

less_naive$table[1:2]

stats <- less_naive$table[[1]]

graph <- tibble(borrower_score = c(0,1)) %>%
    ggplot(aes(borrower_score)) +
    stat_function(
        fun      = dnorm,
        color    = "blue",
        linetype = 1,
        args     = list(
            mean = stats[1, 1],
            sd   = stats[1, 2]
        )
    ) +
    stat_function(
        fun      = dnorm,
        color    = "red",
        linetype = 2,
        args     = list(
            mean = stats[2, 1],
            sd   = stats[2, 2]
        )
    ) +
    labs(y = "probability") +
    theme_classic()

graph
```

## Discriminant Analysis

### A Simple Example

```{r}
loan_lda <- lda(
    outcome ~ borrower_score + payment_inc_ratio,
    data = loan3000
)

loan_lda$scaling

pred <- predict(loan_lda)

head(pred$posterior)
```

#### Figure 5.1

```{r}
pred   <- predict(loan_lda)
lda_df <- bind_cols(
    loan3000,
    prob_default = pred$posterior[,"default"]
)

center    <- 0.5 * (loan_lda$mean[1, ] + loan_lda$mean[2, ])
slope     <- -loan_lda$scaling[1] / loan_lda$scaling[2]
intercept <- center[2] - center[1] * slope

graph <- lda_df %>% 
    ggplot(
        aes(
            x     = borrower_score,
            y     = payment_inc_ratio,
            color = prob_default
        )
    ) +
    geom_point(alpha = 0.6) +
    scale_color_gradientn(
        colors = c("#ca0020", "#f7f7f7", "#0571b0")
    ) +
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0), lim = c(0, 20)) + 
    geom_abline(
        slope     = slope,
        intercept = intercept,
        color     = "darkgreen") +
    theme_classic()

graph
```

## Logistic regression

### Logistic Response Function and Logit

```{r}
logistic_model <- glm(
    outcome ~
        payment_inc_ratio +
        purpose_ +
        home_ +
        emp_len_ +
        borrower_score,
    data   = loan_data,
    family = "binomial"
)

logistic_model

summary(logistic_model)

p   <- seq(from = 0.01, to = 0.99, by = 0.01)
tbl <- tibble(
    p     = p,
    logit = log(p/(1-p)),
    odds  = p/(1-p)
)

graph <- tbl %>% 
    ggplot(aes(p, logit)) +
    geom_line() +
    labs(
        x = "p",
        y = "logit(p)"
    ) +
    theme_classic()

graph
```

### Predicted Values from Logistic Regression

```{r}
pred <- predict(logistic_model)
summary(pred)

prob <- 1/(1 + exp(-pred))
summary(prob)
```

### Interpreting the Coefficients and Odds Ratios

```{r}
graph <- tbl %>% 
    ggplot(aes(logit, odds)) +
    geom_line() +
    labs(
        x = "log(odds ratio)",
        y = "odds ratio"
    ) +
    coord_cartesian(
        xlim = c(0, 5),
        ylim = c(1, 100)
    ) +
    theme_classic()

graph
```

### Logistic regression with splines

```{r}
logistic_gam <- gam(
    outcome ~
        s(payment_inc_ratio) +
        purpose_ +
        home_ +
        emp_len_ +
        s(borrower_score),
    data   = loan_data,
    family = "binomial"
)

logistic_gam
```

### Assessing the Model

```{r}
# 1
terms         <- predict(logistic_gam, type = "terms")
partial_resid <- resid(logistic_gam) + terms

.payment_inc_ratio <- loan_data %>%
    pull("payment_inc_ratio")

.terms <- terms %>%
    as_tibble() %>%
    pull("s(payment_inc_ratio)")

.partial_resid <- partial_resid %>% 
    as_tibble() %>% 
    pull("s(payment_inc_ratio)")

tbl <- tibble(
    .payment_inc_ratio,
    .terms,
    .partial_resid
)

graph <- tbl %>%
    ggplot(aes(.payment_inc_ratio, .partial_resid, solid = FALSE)) +
    geom_point(shape = 46, alpha = 0.4) +
    geom_line(
        aes(.payment_inc_ratio, .terms),
        color     = "red",
        alpha     = 0.5,
        linewidth = 1.5
    ) +
    labs(y = "Partial Residual") +
    coord_cartesian(xlim = c(0, 25)) +
    theme_classic()

graph

# 2

tbl <- tibble(
    .payment_inc_ratio,
    terms = 1 / (1 + exp(-.terms)),
    partial_resid = 1 / (1 + exp(-.partial_resid))
)

graph <- tbl %>%
    ggplot(aes(.payment_inc_ratio, partial_resid, solid = FALSE)) +
    geom_point(shape = 46, alpha = 0.4) +
    geom_line(
        aes(.payment_inc_ratio, terms),
        color     = "red",
        alpha     = 0.5,
        linewidth = 1.5
    ) +
    labs(y = "Partial Residual") +
    coord_cartesian(xlim = c(0, 25)) +
    theme_classic()

graph
```

## Evaluating Classification Models

### Confusion Matrix

```{r}
# Confusion matrix
pred      <- predict(logistic_gam, newdata = loan_data)
pred_y    <- as.numeric(pred > 0)
true_y    <- as.numeric(loan_data$outcome == "default")
true_pos  <- (true_y == 1) & (pred_y == 1)
true_neg  <- (true_y == 0) & (pred_y == 0)
false_pos <- (true_y == 0) & (pred_y == 1)
false_neg <- (true_y == 1) & (pred_y == 0)

conf_mat  <- matrix(
    c(
        sum(true_pos),
        sum(false_pos),
        sum(false_neg),
        sum(true_neg)),
    2,
    2
)

colnames(conf_mat) <- c("Yhat = 1", "Yhat = 0")
rownames(conf_mat) <- c("Y = 1", "Y = 0")

conf_mat
```

### Precision, Recall, and Specificity

```{r}
# Precision
conf_mat[1, 1] / sum(conf_mat[,1])

# Recall
conf_mat[1, 1] / sum(conf_mat[1,])

# Specificity
conf_mat[2, 2] / sum(conf_mat[2,])
```

### ROC Curve

```{r}
idx         <- order(-pred)
recall      <- cumsum(true_y[idx] == 1) / sum(true_y == 1)
specificity <- 
    (sum(true_y == 0) - cumsum(true_y[idx] == 0)) / sum(true_y == 0)
roc_tbl     <- tibble(recall = recall, specificity = specificity)

graph <- roc_tbl %>%
    ggplot(aes(specificity, recall)) +
    geom_line(color = "blue") +
    scale_x_reverse(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    geom_line(
        data     = tibble(x = (0:100) / 100),
        aes(x, 1-x),
        linetype = "dotted",
        color    = "red"
    ) +
    theme_classic()

graph
```

### AUC

```{r}
sum(roc_tbl$recall[-1] * diff(1-roc_tbl$specificity))
head(roc_tbl)

graph <- roc_tbl %>%
    ggplot(aes(specificity)) +
    geom_ribbon(
        aes(ymin = 0, ymax = recall),
        fill = "blue",
        alpha = 0.3
    ) +
    scale_x_reverse(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(y = "recall") +
    theme_classic()

graph
```

## Strategies for Imbalanced Data

### Undersampling

```{r}
full_train_set %>% 
    summarize(
        mean_outcome = mean(outcome=="default")
    ) %>% 
    pull()

full_model <- glm(
    outcome ~
        payment_inc_ratio +
        purpose_ +
        home_ +
        emp_len_+
        dti +
        revol_bal +
        revol_util,
    data   = full_train_set,
    family = "binomial"
)

pred <- predict(full_model)

mean(pred > 0)

full_train_set %>%
    summarize(default_mean = mean(outcome == "default")) %>%
    pull(default_mean) / mean(pred > 0)
```

### Oversampling and Up/Down Weighting

```{r}
full_model <- full_train_set %>%
    mutate(
        wt = if_else(
            outcome == "default",
            1 / mean(outcome == "default"
            ),
            1
        )
    ) %>%
    glm(
        outcome ~
            payment_inc_ratio +
            purpose_ +
            home_ +
            emp_len_+
            dti +
            revol_bal +
            revol_util,
        data   = .,
        weight = wt,
        family = "quasibinomial"
    )

pred <- full_train_set %>%
    add_column(pred = predict(full_model)) %>%
    summarize(mean_pred = mean(pred > 0)) %>%
    pull(mean_pred)

pred
```

### Data Generation

```{r}
# There are a variety of SMOTE implementation available in R. 

# The packages {unbalanced} and {DMwR} are no longer supported, however. So here's a generic example from the {smotefamily}:

library(smotefamily)

data_example = sample_generator(100, ratio = 0.80)
genData      = SMOTE(data_example[,-3], data_example[,3])
genData_2    = SMOTE(data_example[,-3], data_example[,3], K = 7)

genData
genData_2
```

### Exploring the Predictions

```{r}
loan_tree <- rpart(
    outcome ~ borrower_score + payment_inc_ratio,
    data    = loan3000,
    control = rpart.control(cp = 0.005)
)

lda_pred <- tibble(
    borrower_score = c(
        (0 - intercept) / slope,
        (20 - intercept) / slope
    ),
    payment_inc_ratio = c(0, 20),
    method            = rep("LDA", 2)
)

tree_pred <- tibble(
    borrower_score    = c(0.375, 0.375, 0.475, 0.475, 0.575, 0.575),
    payment_inc_ratio = c(0, 4.426,  4.426, 10.42, 10.42, 20),
    method            = rep("Tree", 6)
)

glm0 <- glm(
    outcome ~ (payment_inc_ratio) +  (borrower_score),
    data    = loan3000,
    family  = "binomial"
)

y <- seq(from = 0, to = 20, length = 100)
x <- (-glm0$coefficients[1] - glm0$coefficients[2]*y) / glm0$coefficients[3]

glm0_pred <- tibble(
    borrower_score    = x,
    payment_inc_ratio = y,
    method            = "Logistic"
)

gam1 <- gam(
    outcome ~ s(payment_inc_ratio) +  s(borrower_score),
    data    = loan3000,
    family  = "binomial"
)

gam_fun <- function(x){
    rss <- sum(
        predict(
            gam1,
            newdata = tibble(
                borrower_score    = x,
                payment_inc_ratio = y
            )
        ) ^ 2
    )
}

est_x     <- nlminb(seq(from = 0.33, to = 0.73, length = 100), gam_fun)
gam1_pred <- tibble(
    borrower_score    = est_x$par,
    payment_inc_ratio = y,
    method            = "GAM"
)

loan_fits <- rbind(lda_pred, tree_pred, glm0_pred, gam1_pred)

graph <- loan_fits %>% 
    ggplot(
        aes(
            borrower_score,
            payment_inc_ratio,
            color    = method,
            linetype = method
        )
    ) +
    geom_line(linewidth = 1.5) +
    theme(legend.key.width = unit(2, "cm")) +
    guides(linetype = guide_legend(override.aes = list(size = 1))) +
    theme_classic()

graph
```
